The impact of gas bulk rotation on the Lyman $\alpha$ line

### 2014-06-17 paper/ms.tex
new revised version

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider two types of source distributions: single central and a
homogeneoulsy distributed in the sphere. 
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis the intensity at the line
center and the line width increase with rotational velocity.   
Along the same sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux nor in the escape fraction. 
We provide an analytical expression for the outgoing \lya spectrum of
a rotating cloud which accounts for the velocity and viewing angle
trends observed in the simulations.
\end{abstract}

@@

Ee parametrize the dependency of the line width with  $V_{\rm max}$ as
%
\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the velocity width in the static case and $\lambda$ 
is a positive scalar to be determined as a fit to the data.  
With this test we want to know to what extent the new velocity width can be
expressed as a quadratic sum of the two relevant velocities in the
problem.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f8.pdf}
\end{center}
\caption{Position of the line maxima as a function of maximum
  rotational   velocity $V_{\rm max}$. Continuous (dashed) lines
  correspond to   central (homogeneous) source distributions. A value
  of $x_{\rm     max}=0$ indicates that line becomes single
  peaked. \label{fig:maximumsvsvelocity}}
\end{figure*}

We measure the peak maxima position to quantify the transition from
double into single peak profiles. We do this as a function of the viewing angle
$\theta$ for different rotational velocities. The results are
summarized in Figure  \ref{fig:maximumsvsvelocity}. There are two
interesting features that deserve attention. First, for a viewing angle
parallel to the rotational axis ($\mu\sim 1.0$) the maxima of all
models are similar regardless of the rotational velocity. Second, at a
viewing angle perpendicular to the rotation axis ($\mu\sim 0.0$) a
good fraction of models become single peaked. This feature appears
more frequently for homogeneously distributed sources if all the
other parameters are equal.

@@

\begin{table}
\begin{center}
\begin{tabular}{c ccc ccc}
\hline \hline
Source & $\tau_{H}$ & & & $\ V_{\rm max}$& & \\
Distribution& &  &  & (\kms) & & \\ 
& & 0 & 50 & 100 &200 & 300\\ \hline 
Homogeneous & $10^{5}$& 0.263 & 0.266 &  0.309 &  0.357 &  0.370  \\
            & $10^{6}$ & 0.289 &  0.289 &   0.309 &  0.360 &  0.393 \\
            &$10^{7}$ &  0.227 &  0.229 &  0.231 &  0.254 &  0.281 \\
Central & $10^{5}$ &  0.096 & 0.097 & 0.096 &  0.096 & 0.096 \\
  		&$10^{6}$ & 0.066 & 0.066 &  0.066 &  0.066 &  0.066 \\
 		&$10^{7}$ & 0.015 &  0.016 & 0.016 & 0.016 & 0.015 \\
\hline
\end{tabular}
\caption{
 Escape fraction values for all dusty models. } 
\label{table:escape}
\end{center}
\end{table}

%\begin{figure*}
%\begin{center}
%  \includegraphics[width=0.40\textwidth]{f10.pdf}
%  \includegraphics[width=0.40\textwidth]{f11.pdf}
%\end{center}
%  \caption{Effects of rotation on the escape fraction. Left
%    panel, escape fraction as a function of rotational velocity. All
%    these models have $\tau_{a}=1$. The continuous (dashed) lines
%    correspond to central (homogeneous) models. Right panel, escape
%    fraction as a    function of the product $(a\tau_{\rm
%      H})^{1/3}\tau_{A}$. The    analytic solution for the infinite
%    slab is shown as a continuous line. Different lines correspond
%    to different rotational velocities for the homogeneous models.     
%    \label{fig:escape_fraction}} 
%\end{figure*}

We now estimate the escape fraction $f_{\rm esc}$ for the dusty
models. The main  result is that we do not find any dependence
with the viewing angle but there is a strong dependence with the
rotational velocity for the homogeneous models.

The left panel in Figure \ref{fig:escape_fraction} summarizes the
findings about the dependence with the rotational velocity. The left
panel shows the escape fraction as a function of the rotational
velocity. To construct this Figure we take into account all the
photons regardless of its outgoing direction. We observe that the 
curves for the central source distribution stay flat, while for the
homogeneous case there is a clear rise with rotational velocity.
Rotation has a higher relative impact in the models with low optical
depth. For instance, in models with $\tau_{\rm H}=10^5$, the static
escape fraction is $0.26$ and increases to $0.37$ for $V_{\rm
  max}=300$ \kms. Table \ref{table:escape} lists all the values for
the escape fraction.

In the right panel of Figure \ref{fig:escape_fraction} we put these
results in the context of the analytic solution for the infinite slab
\citep{Neufeld90}. In Neufeld's set-up the analytic solution depends
uniquely on the product $(a\tau_{\rm   H})^{1/3}\tau_{A}$ where
$\tau_{A} = (1 - A)\tau_{a}$, valid only in the limit $a\tau_{\rm
  H}\gg 1$. The dashed lines in  the right panel of Figure
\ref{fig:escape_fraction} show the results for the different
rotational velocities for the homogeneous models. First of all we note
that the escape fraction increases slightly from $\tau_{\rm H}=10^5$ to
$\tau_{\rm H}=10^{6}$, which is counterintuitive as we expected that increasing
$\tau_{\rm H}$ requires more scattering before the photon escapes, which would
increase the probability that photons are destroyed by dust.

However, this naive interpretation is incorrect: increasing $\tau_{\rm
  H}$ from $10^{5}$ to $10^{6}$ causes us to transition from the
'optically thick' to  the 'extremely optically thick'
  regime. In the optically thick regime, Lya  photons can escape in
  'single flight' which corresponds to a scenario in which the
  photon resonantly scatters $10^4-10^5$ times until it is scattered
  into the wing of the line ($x\sim 3-4$). At these frequencies the
  medium is optically thin, and the photons can escape efficiently in
  a single flight. In contrast, in an extremely optically thick medium \ly
  photons escape in a `single excursion' \citep{Adams72}. Here, photons
that are scattered into the wing of the line escape from the medium in
a sequence of wing scattering events.In both cases, \ly photons
resonantly scatter $10^4$-$10^5$ times. Because we keep our clouds the same
size, the mean free path of Lya photons that scatter resonantly is 10
times larger for the case $\tau_{\rm H}=10^5$ than for $\tau_{\rm
  H}=10^6$. If we compute the average distance $D$ travelled by \ly
photons through a medium of size $R$ as a function of line center
optical depth $\tau_{\rm H}$, then we find that during the transition
from optically thick to extremely optically thick the mean traversed
distance $D$ actually decreases slightly.  This decrease is unique to
this transition region, and $D$ generally increases with $\tau_{\rm H}$ at other
values of $\tau_{\rm H}$.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{f14_1.pdf}
  \includegraphics[width=0.48\textwidth]{f14_2.pdf}
  \includegraphics[width=0.48\textwidth]{f14_3.pdf}
  \includegraphics[width=0.48\textwidth]{f14_4.pdf}    
\end{center}
    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
      show the homogeneous (central) source distribution. Left panels 
      corresponds to the static case and right panels correspond to
      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
      number of photons with given values of $N_{\rm scatt}$ and
      $x$. \label{fig:Nscatt2D}. }
\end{figure*}

The number of scatterings affects the escape frequency of a \ly
photon. Studying this quantity can help us in claryfying two results
that we have observed for homogeneous models in the previous sections:
the quick emergence of a single peak and the increase in the escape
fraction.

A possible explanation for these two results is that some photons
close to the surface can escape with a low number of scatterings. This
has a two consequences: the escaping photons do not have time to
scatter away from the line's center and also have less chances to find
an absorbing dust grain.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the cosinus of the
outgoing angle $|\cos\theta|$ and the rotational velocity
$V_{\rm max}$. From the right panel observe that the number of
scatterings and the outgoing angle are independent. This plot
corresponds to the specific case of the central model with $\tau=10^5$ and
$V_{\rm max}=300$\kms, but we have veryfied that this holds for all
models. This allows us to construct statistics using all the outgoing
photons regardless of their value of $\cos\theta$.

The right panel of Figure \ref{fig:Nscatt} shows how the
average number of scatterings decreases for larger rotational
velocities in the case of homogeneous source distributions. For instance, for
$\tau_{\rm H}=10^5$ the average number of scatterings decreases by
$61\%$ at $V_{\rm max}=300\kms$ in comparison to the static case. For
the central source distribution the average number of scatterings
$\langle N_{\rm   scatt}\rangle$ changes less than $0.5\%$.

In order to gain a deeper understanding of these results  we make 2D
histograms for the number of scatterings as a function of the outgoing
dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show these 
results in the case $\tau_{\rm H}=10^5$ for the
static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
results for the homogeneous (central) source distribution. The color
scale is logarithmic in the number of photons at a certain value
$x-N_{\rm scatt}$.

The top right panel of Figure \ref{fig:Nscatt2D} (homogeneous sources,
high rotational velocity) supports our hypothesis about
the photo-sphere in the homogeneous distribution. In this case most of
the photons that left with $x\sim 0$ have escaped with less than $10$
scatterings. This also explains the decrease in the average number of
scatterings observed in Figure \ref{fig:Nscatt}. We have also veryfied
that these photons are indeed emitted closer to the surface than to
the center of the sphere.

However, for a central distribution the situation is quite different
(lower panels in Figure \ref{fig:Nscatt2D}). In this case the number
of scatterings remains high, in the order of the optical depth, but
the two peaks do get closer to each other. The most probable
physical picture is that each scattering, due to the bulk velocity of
the gas, is inefficient in driving the photon outside the line
center.

We now proceed with a more detailed quantification of the number of
scatterings in our simulations. We guide the measurement by the fact
that in static cases, a large value of the optical depth correlates
with a high number of scatterings.  This has been precisely quantified
in the case of static infinite slab. In this model for centrally
emitted sources the average number of scatterings depends only on the
optical depth $\langle N_{\rm  scatt}\rangle=1.612\tau_{\rm   H}$
\citep{Adams72,Harrington73}, for homogeneously distributed sources
$\langle N_{\rm   scatt}\rangle=1.16\tau_{\rm   H}$
\citep{Harrington73}.

We perform a similar quantification of in our speherical setup for the
static spheres. We find that for the central model the number of
scatterings is proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively.
For the homogeneous static sphere we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$, this represents
factors of $(0.66, 0.59, 0.55)$ lower than the centrally  emitted photons.

For the homogeneous rotating sphere at  $V_{\rm{max}}=300\kms$ we find
that $\langle N_{\rm   scatt}\rangle=(0.38, 0.34, 0.42)\tau_{\rm   H}$
which represents   factors of $(0.26, 0.34, 0.46)$ lower than the
corresponding central case.

@@

@@

{\bf Galaxy simulations with gas rotation}. \cite{Verhamme12} studied \ly
line emission in two high resolution simulations of individual
galaxies. The main purpose of their study was to assess the impact of
two different Inter-Stellar Medium (ISM) prescriptions. However, each
simulated galaxy had a disc structure with a clear rotation pattern in
the ISM and inflowing gas from the circum-galactic region. The configuration
had an axial symmetry and they reported a strong dependence of both
the escape fraction and the total line intensity as a function of the
$\theta$ angle. In our results, none of these two quantities has a
dependence either on the inclination angle or the rotational
velocity. We suggest that he effect reported by \cite{Verhamme12} is
consistent with being a consequence of the different hydrogen optical
depth for different viewing angles and not as an effect of the bulk
rotation.

{\bf Single peaked lines}. The presence of single peaked profiles has
been associated to inflow/outflow dynamics
\citep{Verhamme06,DijkstraKramer}. In this paper we show that gas bulk
rotation can also be considered as a probable origin for that
behavior, provided that the observed single peak is highly
symmetric. Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be
considered in the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. This highlights
that in order to interpret the observations of \ly emitting galaxies
it is necessary to consider the possible effect of different
inclination angles along the line of sight.

@@

{\bf Additional uncertainty on the \ly escape fraction}. Study of
high redshift LAEs in numerical simulation often requires the
estimation of the \ly escape fraction in order to compare their
results against observations
\citep{CLARA,Dayal2012,Forero12,Orsi12,Garel2012}. Most of these
models estimate the escape fraction from the column density of dust and
neutral Hydrogen. The results of our simulation suggest that the
rotational velocity has also to be accounted for, having an impact on
the values of the escape fractions by factors of $\sim 2$.

@@

Remarkably, we find two quantities that are invariant with respect
to the viewing angle and the rotational velocity: the integrated flux and the escape
fraction.  The exception is the models with homogeneously distributed
sources, which present an escape fraction dependent on the rotational
velocity, but still independent of the viewing angle. We showed that
these results could be interpreted in terms of the average number of
scatterings for the photons.

@@

\item For a viewing angle with $\mu=0$ the line width increases with rotational
  velocity. This change approximately follows the functional form  ${\rm FWHM}^2 = {\rm FWHM}_{ 0}^2 + (V_{\rm max}/\lambda)^2$, where FWHM$_{0}$ indicates the line
  width for the static case and $\lambda$ is a constant. We have
  determined this constant to be  $\lambda_{\rm c}=0.84 \pm 0.06$ and
  $\lambda_{\rm h}=0.54\pm 0.10$ for the central and homogeneous source
  distributions, respectively.

\item At fixed rotational velocity the line width decreases as $|\mu|$
  increases, i.e. the smallest value of the line width is observed for
  a line of sight parallel to the ration axis. However the lowest
  values are always larger than the values for the static case.

\item The single peaked line emerges for homogeneously distributed sources 
  when the rotational velocity is larger than half FWHM$_0$ and $\mu\sim
  1$.

\item  For central sources we find that the escape fraction does not change
  with rotational velocity nor with the viewing angle. We interpret this
  in terms of the average number of scatterings which does not
  significantly change with rotation. This is due to the  
  fact that the vast majority of scatterings events are resonant, 
  during which the mean free path is very short and the effect of gas
  bulk rotation is not enough to affect \ly photons.

\item For homogeneous sources we find that the  as rotational velocity
  increases, the escape fraction rises by a factor of $\sim 2$ with
  respect to the static situation. We also interpret this change in
  terms of the average number of scatterings. In the second part of \S
  \ref{sec:widthpeak} we show that the average number of scatterings
  decreases a $40\%$ compared to the static case, while a large
  fraction of photons escape with less than $10$ scatterings.

@@

Comparing our results with recent observed LAEs we find that many 
morphological features such as high central line flux, single peak
profiles  could be explained by gas bulk  rotation present in these
LAEs.

@@

Ly$\alpha$ scattering through a an optically thick gas cloud that is
undergoing solid-body rotation (i.e. the angular speed around the
rotation axis is identical for each hydrogen atom) proceeds identical
as in a static cloud. 
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame. 
In the frame of the rotating gas cloud, all atoms are stationary w.r.t
each other, an the scattering process proceeds identical as in the static
case\footnote{An alternative - more quantitative way - to view this is
  to consider the path of an individual photon: say the photon travels
  a distance corresponding to $\tau=1$, scatters by $90$ degrees, then
  travels another distance corresponding to $\tau=1$. It is easy to
  show that the point where the photon ends up is independent of the
  assumed rotation velocity of the cloud.}. 
This explains why the escape fraction of Ly$\alpha$ photon is
independent of the rotation velocity off the cloud.

The spectrum of \lya photons emerging from a rotating gas cloud is
also identical as for the static case in a frame that is co-rotating
with the cloud. However, the surface of cloud now moves in the
lab-frame. Each surface-element on the rotating cloud now has a bulk
velocity w.r.t a distant observer. In order to compute the spectrum
emerging from this rotation cloud, we have to sum over all surface
elements (weighted by their intensity).

@@

\begin{figure*}[h]
\centerline{\includegraphics[width=85mm]{scheme1.pdf}
\includegraphics[width=85mm]{scheme2.pdf}}
\caption[]{Adopted geometry for evaluating the analytic spectrum.}
\label{fig:scheme}
\end{figure*}

@@

Fig~\ref{fig:comparison}shows some examples of analytic vs. full MC
spectra.
The area under the spectra is the same in all cases. 
The left panel shows the results for different rotational velocitiies
in the case of $\tau_{H}=10^7$ and an observer located perpendicular
to the axis of rotation ($\mu=0$ in the notation of the main body of
the paper, $i=0$ in the scheme of Fig~\ref{fig:scheme}). 
The right panel shows the results for different viewing angles in the
case of $\tau_{H}=10^7$ and a rotational velocity of $V_{\rm
  max}=300$\kms.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.48\textwidth]{vary_vel.pdf}
  \includegraphics[width=0.48\textwidth]{vary_angle.pdf}
\end{center}
\caption{
Comparison of the Monte Carlo results agains the analytic
solution. The left panel explores the results of different velocities.
The right panel presents the results for two different observers:
paralel and perpendicular to the rotationa axsi, $\mu=1$ and $\mu=0$
respectively.
\label{fig:comparison} }   
\end{figure*}

>>>>

We present results of radiative transfer calculations to measure the
impact of gas bulk rotation on the morphology of the Lyman $\alpha$ 
emission line in distant galaxies. 
We model a galaxy as a sphere with an homogeneous mixture of dust and
hydrogen at a constant temperature. 
These spheres have a solid-body rotation with maximum velocities in
the range $0-300$ \kms and neutral hydrogen optical depths in the
range $\tau_{\rm H}=10^{5}-10^{7}$. 
We consider two types of source distributions: single central and a
homogeneoulsy distributed in the sphere. 
Our main result is that rotation introduces a dependence of the
line morphology with  viewing angle. 
Observations with a line of sight parallel to the rotation axis yield
line morphologies similar to the static case. 
For lines of sight perpendicular to the rotation axis the intensity at the line
center and the line width increase with rotational velocity.   
Along the same sight, the line becomes single peaked  at rotational
velocities close to half the line width in the static case. 
Notably, we find that rotation does not induce any significant spatial
anisotropy in the integrated line flux, in the escape fraction or the
average number of scatterings. 
We also provide an analytical expression for the outgoing \lya spectrum of
a rotating cloud which accounts for the velocity and viewing angle
trends observed in our Monte Carlo simulations.
\end{abstract}

@@

We parametrize the dependency of the line width with  $V_{\rm max}$ as
%
\begin{equation}
 {\rm FWHM}^2 = {\rm FWHM}_{0}^2 + V_{\rm max}^2/\lambda^2,
\label{eq:fwhm}
\end{equation}
%
where FWHM$_{0}$ is the velocity width in the static case and $\lambda$ 
is a positive scalar to be determined as a fit to the data.  
With this test we want to know to what extent the new velocity width can be
expressed as a quadratic sum of the two relevant velocities in the
problem.

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.95\textwidth]{f8.pdf}
\end{center}
\caption{Position of the line maxima as a function of maximum
  rotational   velocity $V_{\rm max}$. Continuous (dashed) lines
  correspond to   central (homogeneous) source distributions. A value
  of $x_{\rm     max}=0$ indicates that line becomes single
  peaked. \label{fig:maximumvsvelocity}}
\end{figure*}

We measure the peak maxima position, $x_m$, to quantify the transition from
double into single peak profiles. 
In Figure \ref{fig:maximumvsvelocity} we show the dependence of $x_m$ with
the viewing angle parameterized by $|\cos\theta|$ for different
rotational velocities.
There are two interesting features that deserve attention. 
First, for a viewing angle parallel to the rotational axis ($\mu\sim
1.0$) the maxima of all models with the same kind of source
initialization are similar regardless of the rotational velocity.  
Second, at a viewing angle perpendicular to the rotation axis ($\mu\sim 0.0$) a
large fraction of models become single peaked. 
This feature appears more frequently for homogeneously distributed
sources if all the other parameters are equal.

@@

\begin{table}
\begin{center}
\begin{tabular}{c cccccc}
\hline \hline
Source & $\tau_{H}$ & &  $\ V_{\rm max}$& & \\
Distribution& &    & (\kms) & & \\ 
& & 0 & 100 &200 & 300\\ \hline 
Homogeneous & $10^{5}$& 0.263 &  0.263 &  0.263 &  0.263  \\
            & $10^{6}$ & 0.291 &   0.292 &  0.293 &  0.293 \\
            &$10^{7}$ &  0.228 &  0.228 &  0.228 &  0.228 \\
Central & $10^{5}$ &  0.096 & 0.096 &  0.096 & 0.096 \\
  		&$10^{6}$ & 0.066 &  0.066 &  0.066 &  0.066 \\
 		&$10^{7}$ & 0.015 & 0.016 & 0.016 & 0.015 \\
\hline
\end{tabular}
\caption{
 Escape fraction values for all dusty models. } 
\label{table:escape}
\end{center}
\end{table}

We now estimate the escape fraction $f_{\rm esc}$ for the dusty
models. 
The main  result is that we do not find any significant dependence
with either the viewing angle nor the rotational velocity.  
For completeness we list in Table \ref{table:escape} the scape
fraction for all models.

We now put these results in the context of the analytic solution for
the infinite slab\citep{Neufeld90}.   
In Neufeld's set-up the analytic solution depends
uniquely on the product $(a\tau_{\rm   H})^{1/3}\tau_{A}$ where
$\tau_{A} = (1 - A)\tau_{a}$, valid only in the limit $a\tau_{\rm
  H}\gg 1$. 
At fixed values of $\tau_{a}$ the escape fraction monotonically
decreases with increasing values of $\tau_{\rm H}$. 
This expectation holds for the central sources. 
But in the case of homogeneous sources the escape fraction increases
slightly from $\tau_{\rm H}=10^5$ to $\tau_{\rm H}=10^{6}$

The naive interpretation of the analytic solution does not seem to
hold for photons emitted far from the sphere's center. 
We suggest that increasing $\tau_{\rm H}$ from $10^{5}$ to $10^{6}$ causes a
transition from the 'optically thick' to  the 'extremely optically
thick' regime for a noticeable fraction of the photons in the
homogeneous source distribution.

In the optically thick regime, \lya  photons can escape in
'single flight' which corresponds to a scenario in which the
photon resonantly scatters $10^4-10^5$ times until it is scattered
into the wing of the line ($x\sim 3-4$). 
At these frequencies the medium is optically thin, and the photons can
escape efficiently in a single flight. 
In contrast, in an extremely optically thick medium \ly
  photons escape in a `single excursion' \citep{Adams72}. 
Here, photons that are scattered into the wing of the line escape from
the medium in a sequence of wing scattering events. 
In both cases, \ly photons resonantly scatter $10^4$-$10^5$ times. 
Because we keep our clouds the same size, the mean free path of Lya
photons that scatter resonantly is 10 times larger for the case
$\tau_{\rm H}=10^5$ than for $\tau_{\rm H}=10^6$. 
If we compute the average distance $D$ travelled by \ly
photons through a medium of size $R$ as a function of line center
optical depth $\tau_{\rm H}$, then we find that during the transition
from optically thick to extremely optically thick the mean traversed
distance $D$ actually decreases slightly.  
This decrease is unique to this transition region, and $D$ generally
increases with $\tau_{\rm H}$ at other values of $\tau_{\rm H}$.

@@

\begin{figure*}
\begin{center}
    \includegraphics[width=0.45\textwidth]{f12.pdf}
    \includegraphics[width=0.45\textwidth]{f13.pdf}
\end{center}
\caption{Logarithm of the average number of scatterings as function of
  $\mu$ (left panel) and the maximum rotational velocity $V_{\rm
max}$. The left panel shows the behaviour for $\tau=10^{5}$ and
  $V_{max}=300$\kms as a function of $|\cos\theta|$; the independence of
  $N_{\rm scatt}$ with $\mu$ is present in all models.
  In the right panel the continuous (dashed) lines represent an
  central (homogeneous) distribution of sources; there is a clear
  decrease in the number of scatterings for the homogenously
  distributed sources.
\label{fig:Nscatt} }
\end{figure*}

%\begin{figure*}
%\begin{center}
%  \includegraphics[width=0.48\textwidth]{f14_1.pdf}
%  \includegraphics[width=0.48\textwidth]{f14_2.pdf}
%  \includegraphics[width=0.48\textwidth]{f14_3.pdf}
%  \includegraphics[width=0.48\textwidth]{f14_4.pdf}    
%\end{center}
%    \caption{2D histogram of $N_{\rm scatt}$ vs $x$. The upper (lower) panels
%      show the homogeneous (central) source distribution. Left panels 
%      corresponds to the static case and right panels correspond to
%      $V_{\rm max}=300\kms$. The color scale is logarithmic on the
%      number of photons with given values of $N_{\rm scatt}$ and
%      $x$. \label{fig:Nscatt2D}. }
%\end{figure*}

The number of scatterings affects the escape frequency of a \ly
photon. 
Studying this quantity can help us in claryfying the independence of
the integrated flux and the escape fraction with respect to the
rotational velocity.

In Figure \ref{fig:Nscatt} we show the average number of scatterings
$\langle N_{\rm scatt}\rangle$ as a function of the cosinus of the
outgoing angle $|\cos\theta|$ and the rotational velocity
$V_{\rm max}$. 
From the right panel observe that the number of scatterings and the
outgoing angle are independent. 
This plot corresponds to the specific case of the central model with
$\tau=10^5$ and $V_{\rm max}=300$\kms, but we have veryfied that this
holds for all models.

The right panel of Figure \ref{fig:Nscatt} shows how the average
number of scatterings is also independent from the rotational
velocity. 
The lower number of average scatterings in the homogeneous source
distribution is due to a purely geometrical effect.
Photons emitted close to the surface go through less scatterings
before escaping.

%In order to gain a deeper understanding of these results  we make 2D
%histograms for the number of scatterings as a function of the outgoing
%dimensionless frequency $x$. In Figure \ref{fig:Nscatt2D} we show these 
%results in the case $\tau_{\rm H}=10^5$ for the
%static case and $V_{\rm max}=300$\kms. The upper (lower) panels show the
%results for the homogeneous (central) source distribution. The color
%scale is logarithmic in the number of photons at a certain value
%$x-N_{\rm scatt}$.

%The top right panel of Figure \ref{fig:Nscatt2D} (homogeneous sources,
%high rotational velocity) supports our hypothesis about
%the photo-sphere in the homogeneous distribution. In this case most of
%the photons that left with $x\sim 0$ have escaped with less than $10$
%scatterings. This also explains the decrease in the average number of
%scatterings observed in Figure \ref{fig:Nscatt}. We have also veryfied
%that these photons are indeed emitted closer to the surface than to
%the center of the sphere.

%However, for a central distribution the situation is quite different
%(lower panels in Figure \ref{fig:Nscatt2D}). In this case the number
%of scatterings remains high, in the order of the optical depth, but
%the two peaks do get closer to each other. The most probable
%physical picture is that each scattering, due to the bulk velocity of
%the gas, is inefficient in driving the photon outside the line
%center.

In static configurations it is expected that the optical depth correlates number of
scatterings.  
This has been precisely quantified in the case of static infinite
slab. 
In that model for centrally emitted sources the average number of
scatterings depends only on the optical depth $\langle N_{\rm
  scatt}\rangle=1.612\tau_{\rm   H}$ \citep{Adams72,Harrington73}, for
homogeneously distributed sources $\langle N_{\rm
  scatt}\rangle=1.16\tau_{\rm   H}$ \citep{Harrington73}.

In our case we find that for the central model the number of
scatterings is proportional to the optical depth, with $\langle N_{\rm
  scatt}\rangle= (1.50, 1.00, 0.92)\tau_{\rm   H}$ for optical depth
values of $\tau_{\rm H} = (10^{5}, 10^{6}, 10^{7})$ respectively.
For the homogeneous sources we find that $\langle N_{\rm
  scatt}\rangle= (0.99, 0.59, 0.51)\tau_{\rm   H}$.

@@

\subsection{Towards an analytical description}

There are three key results that allows us to build an analyitical description to
first order of our results: the independence of the integrated flux,
the number of scatterings and the escape fraction with the rotational
velocity and the viewing angle.

One way to understand these results is that actually a rotating sphere
is not very different from a static one. We can think that in the
frame of the rotating gas cloud all atoms are stationary with respect
to each other, making the scattering process proceeed identical as in
the static case.  
While scattering events off atoms within the rotating cloud impart
Doppler boosts on the Ly$\alpha$ photon, these Doppler boost are only
there in the lab-frame.  
In the frame of the rotating gas cloud, all atoms are stationary with
respect to each othe and the scattering process proceeds identical as
in the static case.

An alternative - more quantitative way - to view this is   to consider
the path of an individual photon: say the photon travels  a distance
corresponding to $\tau=1$, scatters by $90$ degrees, then travels
another distance corresponding to $\tau=1$. 
It is easy to show that the point where the photon ends up is
independent of the assumed rotation velocity of the cloud.

The spectrum of \lya photons emerging from a rotating gas cloud is
also identical as for the static case in a frame that is co-rotating
with the cloud. However, the surface of cloud now moves in the
lab-frame. Each surface-element on the rotating cloud now has a bulk
velocity with respect to a distant observer. In order to compute the
spectrum one can integrate over all the surface elements in the
sphere with their corresponding shift in velocity and an additional
weight by the surface intensity.

Fig~\ref{fig:comparison} shows some examples of analytic vs. full MC
spectra using this approach (the implementation details are in the Appendix).
The left panel shows the results for different rotational velocitiies
in the case of $\tau_{H}=10^7$ and an observer located perpendicular
to the axis of rotation ($i=0$ in the scheme of Fig~\ref{fig:scheme}
in the Appendix). 
The right panel shows the results for different viewing angles in the
case of $\tau_{H}=10^7$ and a rotational velocity of $V_{\rm
  max}=300$\kms.

\begin{figure*}
\begin{center}
  \includegraphics[width=0.49\textwidth]{fig10a.pdf}
  \includegraphics[width=0.49\textwidth]{fig10b.pdf}
\end{center}
\caption{
Comparison of the Monte Carlo results agains the analytic
solution. The left panel explores the results of different velocities.
The right panel presents the results for two different observers:
paralel and perpendicular to the rotationa axsi, $\mu=1$ and $\mu=0$
respectively.
\label{fig:comparison} }   
\end{figure*}

\subsection{Impact on the interpretation of simulated and
  observational data}

@@

{\bf Galaxy simulations with gas rotation}. \cite{Verhamme12} studied \ly
line emission in two high resolution simulations of individual
galaxies. 
The main purpose of their study was to assess the impact of two
different Inter-Stellar Medium (ISM) prescriptions. 
However, each simulated galaxy had a disc structure with a clear rotation pattern in
the ISM and inflowing gas from the circum-galactic region. 
The configuration had an axial symmetry and they reported a strong dependence of both
the escape fraction and the total line intensity as a function of the
$\theta$ angle. 
From our study, none of these two quantities has a dependence either
on the inclination angle or the rotational velocity. 
We suggest that he effect reported by \cite{Verhamme12} is
consistent with being a consequence of the different hydrogen optical
depth for different viewing angles and not as an effect of the bulk
rotation.

{\bf Zero impact on the \ly escape fraction}. Study of
high redshift LAEs in numerical simulation often requires the
estimation of the \ly escape fraction in order to compare their
results against observations
\citep{CLARA,Dayal2012,Forero12,Orsi12,Garel2012}. Most of these
models estimate the escape fraction from the column density of dust and
neutral Hydrogen. The results of our simulation suggest that the
rotational velocity does not induce additional uncertainties in those
estimates.

{\bf Single peaked lines}. The presence of single peaked profiles has
been associated to inflow/outflow dynamics
\citep{Verhamme06,DijkstraKramer}. 
Gas bulk rotation can also be considered as a probable origin for that
behavior, provided that the observed single peak is highly
symmetric. 
Similarly, in the case of double peaked lines with a high
level of flux at the line center, rotation also deserves to be
considered in the pool of possible bulk flows responsible for that feature,
specially if the two peaks have similar intensities. 
This highlights
that in order to interpret the observations of \ly emitting galaxies
it is necessary to consider the possible effect of different
inclination angles along the line of sight.

@@

@@

Remarkably, we find three quantities that are invariant with respect
to the viewing angle and the rotational velocity: the integrated flux,
the escape fraction and the average number of scatterings.
These results helped us to construct the outgoing spectra of a
rotating sphere as a supeposition of spectra coming from a static
configuration. This description is useful to describe the main
quantitative features of the Monte Carlo simulations.

@@

\item For a viewing angle with $\mu=0$ the line width increases with rotational
  velocity. This change approximately follows the functional form  ${\rm FWHM}^2 = {\rm FWHM}_{ 0}^2 + (V_{\rm max}/\lambda)^2$, where FWHM$_{0}$ indicates the line
  width for the static case and $\lambda$ is a constant. We have
  determined this constant to be  $\lambda_{\rm c}=0.83 \pm 0.06$ and
  $\lambda_{\rm h}=0.82\pm 0.05$ for the central and homogeneous source
  distributions, respectively.

\item At fixed rotational velocity the line width decreases as $|\mu|$
  increases, i.e. the smallest value of the line width is observed for
  a line of sight parallel to the ration axis.

\item The single peaked line emerges at viewing angles $\mu\sim 1$ for
  when the rotational velocity is close to than half the FWHM$_0$.

\item  The escape fraction does not change with rotational velocity
  nor with the viewing angle. We interpret this
  in terms of the average number of scatterings which does not
  significantly change with rotation. This is due to the  
  fact that the vast majority of scatterings events are resonant, 
  during which the mean free path is very short and the effect of gas
  bulk rotation is not enough to affect \ly photons.

@@

Comparing our results with recent observed LAEs we find that 
morphological features such as high central line flux, single peak
profiles  could be explained by gas bulk  rotation present in these
LAEs.

@@

Ly$\alpha$ scattering through a an optically thick gas cloud that is
undergoing solid-body rotation (i.e. the angular speed around the
rotation axis is identical for each hydrogen atom) proceeds identical
as in a static cloud. 
In order to compute the spectrum emerging from this rotation cloud, we
have to sum over all surface elements weighted by their intensity.

@@

\begin{figure*}[h]
\centerline{\includegraphics[width=85mm]{fig11a.pdf}
\includegraphics[width=85mm]{fig11b.pdf}}
\caption[]{Adopted geometry for evaluating the analytic spectrum.}
\label{fig:scheme}
\end{figure*}

@@
